---
title: "STAT3D Report"
author: "STAT3D"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    embed-resources: true
    fig-width: 8
    fig-height: 6
execute:
  echo: false
  eval: true
  timeout: 300
  startup_timeout: 300
params:
  results_dir: "results"
  config_file: "run_config.yaml"
---

```{r}
results_dir <- normalizePath(params$results_dir, winslash = "/", mustWork = FALSE)

p <- function(rel_path) {
  file.path(results_dir, rel_path)
}

rel_link <- function(path) {
  sub(paste0("^", results_dir, "/?"), "", normalizePath(path, winslash = "/", mustWork = FALSE))
}

link_or_missing <- function(path) {
  if (file.exists(path) || dir.exists(path)) {
    sprintf("[%s](%s)", basename(path), rel_link(path))
  } else {
    sprintf("(missing) %s", basename(path))
  }
}

markdown_list <- function(items) {
  cat(paste0("- ", items, collapse = "\n"))
}

embed_plot <- function(path, height = 520) {
  if (!file.exists(path)) {
    return(sprintf("<em>Missing:</em> %s", basename(path)))
  }
  png_path <- sub("\\.pdf$", ".png", path)
  if (file.exists(png_path)) {
    return(sprintf(
      "<img src=\"%s\" alt=\"%s\" style=\"max-width:100%%; height:auto; border:1px solid #eee;\"/>",
      rel_link(png_path), basename(png_path)
    ))
  }
  src <- rel_link(path)
  sprintf(
    "<object data=\"%s\" type=\"application/pdf\" width=\"100%%\" height=\"%s\" style=\"border:1px solid #eee;\">PDF: <a href=\"%s\">%s</a></object>",
    src, height, src, basename(path)
  )
}
```

## Overview

Results directory: `r results_dir`

## Parameters

```{r}
if (file.exists(params$config_file)) {
  # Requires r-yaml or just use base parsing for simple structures
  # We assume yaml is installed as standard with rmarkdown
  cfg <- yaml::read_yaml(params$config_file)
  
  # Helper to format values
  fmt_val <- function(x) {
    if (is.null(x)) return("NULL")
    if (is.logical(x)) return(ifelse(x, "TRUE", "FALSE"))
    if (is.list(x) || length(x) > 1) return(paste(x, collapse = ", "))
    as.character(x)
  }
  
  df_cfg <- data.frame(
    Parameter = names(cfg),
    Value = vapply(cfg, fmt_val, character(1)),
    stringsAsFactors = FALSE
  )
  
  # Sort alphabetically
  df_cfg <- df_cfg[order(df_cfg$Parameter), ]
  
  knitr::kable(df_cfg, row.names = FALSE)
} else {
  cat("Configuration file not found.")
}
```

## Effective Analysis Logic

This section details how key parameters were determined (User-defined vs. Auto-calculated).

```{r}
# Function to safely read single value CSVs
read_val <- function(fname, col_name) {
  fpath <- p(fname)
  if (file.exists(fpath)) {
    df <- read.csv(fpath)
    if (nrow(df) > 0 && col_name %in% names(df)) {
      return(round(df[[col_name]][1], 2))
    }
  }
  return(NA)
}

# --- 1. Slice Selection Logic ---
slice_logic <- "Unknown"
slice_details <- ""

# Check Config priorities (consistent with pipeline logic)
if (!is.null(cfg$Z_1)) {
  slice_logic <- "Manual Selection (Config)"
  # Collect all Z_n keys
  z_keys <- grep("^Z_[0-9]+$", names(cfg), value = TRUE)
  z_vals <- unlist(cfg[z_keys])
  slice_details <- paste("Slices:", paste(z_vals, collapse = ", "))
} else if (!is.null(cfg$Z_STACK) && cfg$Z_STACK == "ALL") {
  slice_logic <- "Max Projection (All Slices)"
  slice_details <- "Using entire Z-Stack"
} else {
  # Default: Laplacian Based
  slice_logic <- "Auto-Selection (Laplacian)"
  lap_file <- p("preprocessing/Laplacian_score.csv")
  if (file.exists(lap_file)) {
    lap_df <- read.csv(lap_file)
    # Pipeline default logic: Sort by score desc, take top 2 (or top 1 if HIGHEST)
    if (!is.null(cfg$Laplacian) && cfg$Laplacian == "HIGHEST") {
       top_slices <- lap_df[order(lap_df$Sharpness_score, decreasing = TRUE), ][1, "Z_slice"]
       slice_details <- paste("Top 1 Slice based on Sharpness:", top_slices)
    } else {
       # Default is usually top 2
       top_slices <- lap_df[order(lap_df$Sharpness_score, decreasing = TRUE), ][1:min(2, nrow(lap_df)), "Z_slice"]
       slice_details <- paste("Top 2 Slices based on Sharpness:", paste(top_slices, collapse = ", "))
    }
  }
}

# --- 2. Nuclei Diameter ---
nuclei_logic <- "Unknown"
nuclei_val <- NA

if (!is.null(cfg$nuclei_diameter)) {
  nuclei_logic <- "User Defined (Config)"
  nuclei_val <- cfg$nuclei_diameter
} else {
  nuclei_logic <- "Auto-Calculated"
  nuclei_val <- read_val("segmentation/nuclei_diameter.csv", "Average_Diameter")
}

# --- 3. Nuclear Expansion ---
exp_logic <- "Unknown"
exp_val <- NA
# Check for NUCLEAR_EXPANSION_SET (as per your description)
if (!is.null(cfg$NUCLEAR_EXPANSION_SET)) {
  exp_logic <- "User Defined (Config)"
  exp_val <- cfg$NUCLEAR_EXPANSION_SET
} else {
  exp_logic <- "Auto-Calculated"
  exp_val <- read_val("segmentation/average_nuclear_expansion.csv", "Average_Nuclear_Expansion")
}

# Create Summary Table
df_logic <- data.frame(
  Parameter = c("Z-Slice Selection", "Nuclei Diameter", "Nuclear Expansion"),
  Method = c(slice_logic, nuclei_logic, exp_logic),
  Value = c(slice_details, paste(nuclei_val, "pixels"), paste(exp_val, "pixels")),
  stringsAsFactors = FALSE
)

knitr::kable(df_logic, caption = "Key logical decisions made by the pipeline.")
```

## Preprocessing

```{r, results="asis"}
preprocessing_files <- list(
  c("Laplacian score", p("preprocessing/Laplacian_score.csv")),
  c("Z-slice measurements", p("preprocessing/z_slice_measurements.csv")),
  c("Processed image", p("preprocessing/"))
)
markdown_list(vapply(preprocessing_files, function(item) {
  paste0(item[1], ": ", link_or_missing(item[2]))
}, character(1)))
```

```{r}
laplacian_csv <- p("preprocessing/Laplacian_score.csv")
if (file.exists(laplacian_csv)) {
  head(read.csv(laplacian_csv), 10)
}
```

## Segmentation

```{r, results="asis"}
segmentation_files <- list(
  c("QuPath measurements", p("segmentation/QuPath_measurements.csv")),
  c("Nuclei diameter", p("segmentation/nuclei_diameter.csv")),
  c("Average min distance", p("segmentation/average_min_distance.csv")),
  c("Average nuclear expansion", p("segmentation/average_nuclear_expansion.csv")),
  c("Segmentation mask (NPY)", p("segmentation/"))
)
markdown_list(vapply(segmentation_files, function(item) {
  paste0(item[1], ": ", link_or_missing(item[2]))
}, character(1)))
```

```{r}
expansion_csv <- p("segmentation/average_nuclear_expansion.csv")
if (file.exists(expansion_csv)) {
  head(read.csv(expansion_csv), 10)
}
```

## Counts

```{r, results="asis"}
count_files <- list(
  p("counts/matrix.mtx.gz"),
  p("counts/features.tsv.gz"),
  p("counts/barcodes.tsv.gz")
)
markdown_list(vapply(count_files, link_or_missing, character(1)))
```

```{r}
features_tsv <- p("counts/features.tsv.gz")
if (file.exists(features_tsv)) {
  head(read.delim(features_tsv, header = FALSE), 10)
}
```

```{r}
barcodes_tsv <- p("counts/barcodes.tsv.gz")
if (file.exists(barcodes_tsv)) {
  head(read.delim(barcodes_tsv, header = FALSE), 10)
}
```

## Analysis

```{r, results="asis"}
analysis_files <- list(
  p("analysis/sp_obj.rds"),
  p("analysis/singler.rds")
)
markdown_list(vapply(analysis_files, link_or_missing, character(1)))
```

## Plots

```{r, results="asis"}
plot_files <- list(
  p("plots/QC.pdf"),
  p("plots/spatialobj_plot.pdf"),
  p("plots/SingleR_predictions_QC.pdf"),
  p("plots/Spatial_SingleR.pdf"),
  p("plots/UMAP_SingleR.pdf")
)
markdown_list(vapply(plot_files, link_or_missing, character(1)))
```

```{r, results="asis"}
cat(paste0(
  "<h3>Plot previews</h3>\n",
  paste(vapply(plot_files, embed_plot, character(1)), collapse = "\n\n")
))
```

## Benchmarks

```{r}
bench_file <- p("benchmarks/workflow_resource_usage.txt")

if (file.exists(bench_file)) {
  lines <- readLines(bench_file)
  lines <- lines[grepl("=", lines)]
  parts <- strsplit(lines, "=", fixed = TRUE)
  
  key <- vapply(parts, `[`, 1, FUN.VALUE = character(1))
  val <- vapply(parts, `[`, 2, FUN.VALUE = character(1))
  
  df_bench <- data.frame(Metric = key, Value = val, stringsAsFactors = FALSE)
  
  # Remove max_rss_kb if max_rss_mb exists, to show only one memory metric
  if ("max_rss_mb" %in% df_bench$Metric) {
    df_bench <- df_bench[df_bench$Metric != "max_rss_kb", ]
    df_bench$Metric[df_bench$Metric == "max_rss_mb"] <- "Max RSS (MB)"
  }
  
  # Format other keys for better readability
  df_bench$Metric[df_bench$Metric == "start_ts"] <- "Start Time"
  df_bench$Metric[df_bench$Metric == "end_ts"] <- "End Time"
  df_bench$Metric[df_bench$Metric == "elapsed_seconds"] <- "Duration (s)"
  df_bench$Metric[df_bench$Metric == "exit_code"] <- "Exit Code"
  
  knitr::kable(df_bench, row.names = FALSE)
} else {
  cat("Benchmarks file not found.")
}
```

## Notes

This report is a lightweight summary of available outputs. We will expand it as we finalize the list of key deliverables.